---
title: "Final Project"
output: html_notebook
---

```{r}
# PROLOG   ####################################################################'


# PROJECT: FINAL PROJECT
# PURPOSE: ANALYZE THE ASSOCIATION BETWEEN CONVENTIONAL CIGARETTE SMOKING AND VAPING.
# DATA:    YRBSS 2017
# AUTHORS: RUTH KATUMBA, VIANCA CUEVAS SOULETTE AND DAMILOLA SHOBIYE.
# CREATED: 04/10/19
# LATEST:  04/23/19
# NOTES:   for Advanced Data Analysis class 


# PROLOG  ######################################################################
```

## STEP 1: Load the libraries we'll need. 

```{r}
library(foreign) #for reading spss (read.spss()), stata (read.dta()) and xpt (read.xport()) files
library(haven) #for reading SPSS (read_sav()), stata (read_dta()), SAS(read_sas()) and xpt (read_xpt()) files
library(readr) #for reading csv file (read_csv())
library(knitr) #for creating nicer tables
library(plyr)
library(ggplot2)
library(tidyverse)
library(car)
library(DescTools)
library(lmtest) #for LR test
library(VIM) 
library(mice) 
library(lattice)
library(odds.n.ends)
```


## STEP 2: Load the data.

```{r}
FinalProject_csv <- read_csv("https://raw.githubusercontent.com/Dammiesho/ADA-Group-Project/master/Variable_Interest3_YRBS_17%20FINAL.csv?token=ALCYJ5SRQMSJECNTPKNHQQS4X7U5I")
```


## STEP 3: Rename the variables. 

```{r}
names(FinalProject_csv)<-c("age", "sex", "grade_level", "race", "cig_use", "vap_yn", "vap_cat", "bingedrink", "mari_use", "coke_use", "heroin_use", "meth_use", "ecstasy", "gender", "grades") #renames variables in order of appearance
kable(summary(FinalProject_csv)) #creates 'nice' looking table of summary stats for each variable. Alternatively, use varlist<-colnames (C1Survey_csv)
```


## STEP 4: Classify the variables as factors. 

```{r}
# Classify categorical variables as factor variables
FinalProject_csv$age<-as.factor(as.character(FinalProject_csv$age))
FinalProject_csv$sex<-as.factor(as.character(FinalProject_csv$sex))
FinalProject_csv$grade_level<-as.factor(as.character(FinalProject_csv$grade_level))
FinalProject_csv$race<-as.factor(as.character(FinalProject_csv$race))
FinalProject_csv$cig_use<-as.factor(as.character(FinalProject_csv$cig_use))
FinalProject_csv$vap_yn<-as.factor(as.character(FinalProject_csv$vap_yn))
FinalProject_csv$vap_cat<-as.factor(as.character(FinalProject_csv$vap_cat))
FinalProject_csv$bingedrink<-as.factor(as.character(FinalProject_csv$bingedrink))
FinalProject_csv$mari_use<-as.factor(as.character(FinalProject_csv$mari_use))
FinalProject_csv$coke_use<-as.factor(as.character(FinalProject_csv$coke_use))
FinalProject_csv$heroin_use<-as.factor(as.character(FinalProject_csv$heroin_use))
FinalProject_csv$meth_use<-as.factor(as.character(FinalProject_csv$meth_use))
FinalProject_csv$ecstasy<-as.factor(as.character(FinalProject_csv$ecstasy))
FinalProject_csv$gender<-as.factor(as.character(FinalProject_csv$gender))
FinalProject_csv$grades<-as.factor(as.character(FinalProject_csv$grades))

# Check that you have the right number of factor variables
table(sapply(FinalProject_csv, class))

# Check one variable to comfirm missing data
table(FinalProject_csv$age)
```


## STEP 5: Rename categories.

```{r}
# Rename categories within variables
FinalProject_csv$age[FinalProject_csv$age <= 1] <- 1 
FinalProject_csv$race[FinalProject_csv$race <= 3] <- 2

## Recode age variable 
table(FinalProject_csv$age)
FinalProject_csv$agef<-factor(FinalProject_csv$age, levels = c(1,2,3), labels = c("12 years old or younger", "13 to 17 year-olds", "18 years old or older"))
table(FinalProject_csv$agef)

table(FinalProject_csv$sex)
FinalProject_csv$sexf<-factor(FinalProject_csv$sex, levels = c(1,2), labels = c("Female", "Male"))
table(FinalProject_csv$sexf)

table(FinalProject_csv$grade_level)
FinalProject_csv$grade_levelf<-factor(FinalProject_csv$grade_level, levels = c(1,2,3,4,5), labels = c("9th grade", "10th grade", "11th grade", "12th grade", "Unknown"))
table(FinalProject_csv$grade_levelf)

## Recode race into 3 categories then continue renaming
FinalProject_csv$race[FinalProject_csv$race <= 1] <- 1 
FinalProject_csv$race[FinalProject_csv$race <= 3] <- 2
FinalProject_csv$race[FinalProject_csv$race <= 2 | FinalProject_csv$race <= 4 | FinalProject_csv$race <= 5 | FinalProject_csv$race <= 2 | FinalProject_csv$race <= "F" | FinalProject_csv$race <= "1F" | FinalProject_csv$race <= "5 F" | FinalProject_csv$race <= "1 G" | FinalProject_csv$race <= "H" | FinalProject_csv$race <= "2  H"] <- 3 

table(FinalProject_csv$race)
FinalProject_csv$racef<-factor(FinalProject_csv$race, levels = c(1,2,3), labels = c("White", "Black or African American", "Other races"))
table(FinalProject_csv$racef)

table(FinalProject_csv$cig_use)
FinalProject_csv$cig_usef<-factor(FinalProject_csv$cig_use, levels = c(1,2,3), labels = c("Non-smoker", "Non-daily/Intermittent smoker", "Daily smoker" ))
table(FinalProject_csv$cig_usef)

table(FinalProject_csv$vap_yn)
FinalProject_csv$vap_ynf<-factor(FinalProject_csv$vap_yn, levels = c(0,1), labels = c("No", "Yes"))
table(FinalProject_csv$vap_ynf)

table(FinalProject_csv$bingedrink)
FinalProject_csv$bingedrinkf<-factor(FinalProject_csv$bingedrink, levels = c(1,2,3), labels = c("Non-binge drinker", "Intermittent binge drinker", "Regular binge drinker"))
table(FinalProject_csv$bingedrinkf)

table(FinalProject_csv$mari_use)
FinalProject_csv$mari_usef<-factor(FinalProject_csv$mari_use, levels = c(1,2,3,4), labels = c("No marijuana use", "Light marijuana use", "Regular marijuana use", "Chronic marijuana use" ))
table(FinalProject_csv$mari_usef)

table(FinalProject_csv$coke_use)
FinalProject_csv$coke_usef<-factor(FinalProject_csv$coke_use, levels = c(1,2), labels = c("No", "Yes"))
table(FinalProject_csv$coke_usef)

table(FinalProject_csv$heroin_use)
FinalProject_csv$heroin_usef<-factor(FinalProject_csv$heroin_use, levels = c(1,2), labels = c("No", "Yes"))
table(FinalProject_csv$heroin_usef)

table(FinalProject_csv$meth_use)
FinalProject_csv$meth_usef<-factor(FinalProject_csv$meth_use, levels = c(1,2), labels = c("No", "Yes"))
table(FinalProject_csv$meth_usef)

table(FinalProject_csv$ecstasy)
FinalProject_csv$ecstasyf<-factor(FinalProject_csv$ecstasy, levels = c(1,2), labels = c("No", "Yes"))
table(FinalProject_csv$ecstasyf)

table(FinalProject_csv$gender)
FinalProject_csv$genderf<-factor(FinalProject_csv$gender, levels = c(1,2,3,4), labels = c("Heterosexual", "Gay or Lesbian", "Bisexual", "Not sure"))
table(FinalProject_csv$genderf)

## Recode grades into 4 categories
FinalProject_csv$grades[FinalProject_csv$grades <= 1] <- 1 
FinalProject_csv$grades[FinalProject_csv$grades <= 2] <- 2
FinalProject_csv$grades[FinalProject_csv$grades <= 3] <- 3
FinalProject_csv$grades[FinalProject_csv$grades <= 4 | FinalProject_csv$grades <= 5 | FinalProject_csv$grades <= 6 | FinalProject_csv$grades <= 7 ] <- 4 

table(FinalProject_csv$grades)
FinalProject_csv$gradesf<-factor(FinalProject_csv$grades, levels = c(1,2,3,4), labels = c("Mostly A's", "Mostly B's", "Mostly C's", "Other grades"))
table(FinalProject_csv$gradesf)
```


## STEP 6: Execute imputation to deal with missing data.
```{r}
# Data Imputation
imp<-mice(FinalProject_csv, m=5, seed = 1000) #m = number of imputations, 5 is the default. If you don't include the m argument, 5 imputations will be done.

##look at imputation details. 
imp
```


## STEP 7: Run diagnostics for imputation model: Are the imputed values plausible? 
```{r}
#check age
imp$imp$age #imp$imp refers to dataframe and list called imp within dataframe that has 15 items 
#check sex
imp$imp$sex
#check grade_level
imp$imp$grade_level
#check race
imp$imp$race
#check cig-use
imp$imp$cig_use 
#check vap_yn
imp$imp$vap_yn
#check vap_cat
imp$imp$vap_cat
#check bingedrink
imp$imp$bingedrink
#check mari_use
imp$imp$mari_use 
#check coke_use
imp$imp$coke_use
#check heroin_use
imp$imp$heroin_use
#check meth_use
imp$imp$meth_use
#check ecstasy
imp$imp$ecstasy 
#check gender
imp$imp$gender
#check grades
imp$imp$grades
#the rows contain the imputed values and the columns are the multiple imputations
```


## STEP 7: Obtain the first and second complete datasets for review.
```{r}
complete(imp)
complete(imp,2)
```


## STEP 8: Create a complete dataset.
```{r}
Complete <- na.omit(FinalProject_csv)
```

## Calculate the proportions for Vaping status by sample characteristics.

```{r}
vap_cig0<-table(Complete$cig_usef, Complete$vap_ynf) #Calculating counts
vap_cig0
vap_cig <- prop.table (vap_cig0) #Calculating proportions. 
vap_cig

vap_sex0<-table(Complete$sexf, Complete$vap_ynf) #Calculating counts
vap_sex0
vap_sex <- prop.table (vap_sex0) #Creating table 
vap_sex

vap_race0<-table(Complete$racef, Complete$vap_ynf) #Calculating counts
vap_race0
vap_race <- prop.table (vap_race0) #Creating table 
vap_race


vap_grade0 <-table(Complete$grade_levelf, Complete$vap_ynf) #Calculating counts
vap_grade0
vap_grade <- prop.table (vap_grade0) #Creating table 
vap_grade


vap_gender0 <- table(Complete$genderf, Complete$vap_ynf)
vap_gender0
vap_gender <- prop.table (vap_gender0) #Creating table 
vap_gender



vap_binge0 <- table(Complete$bingedrinkf, Complete$vap_ynf)
vap_binge0
vap_binge <- prop.table (vap_binge0) #Creating table 
vap_binge


vap_mari0 <- table(Complete$mari_usef, Complete$vap_ynf)
vap_mari0
vap_mari <- prop.table (vap_mari0) #Creating table 
vap_mari


vap_coke0 <- table(Complete$coke_usef, Complete$vap_ynf)
vap_coke0
vap_coke <- prop.table (vap_coke0) #Creating table 
vap_coke


vap_ecstasy0 <- table(Complete$ecstasyf, Complete$vap_ynf)
vap_ecstasy0
vap_ecstasy <- prop.table (vap_ecstasy0) #Creating table 
vap_ecstasy


vap_meth0 <- table(Complete$meth_usef, Complete$vap_ynf)
vap_meth0
vap_meth <- prop.table (vap_meth0) #Creating table 
vap_meth


vap_grades0 <- table(Complete$gradesf, Complete$vap_ynf)
vap_grades0
vap_grades <- prop.table (vap_grades0) #Creating table 
vap_grades


vap_her0 <- table(Complete$heroin_usef, Complete$vap_ynf)
vap_her0
vap_her <- prop.table (vap_her0) #Creating table 
vap_her
```


## STEP 9: Run descriptive analysis for complete dataset.
```{r}
install.packages("Hmisc") #install package to run descriptive of each variable
library(Hmisc)

describe(Complete)
```


## STEP 10: Run Univariate Analysis. 

```{r}
#Association with the outcome 
VapingCiglog <- glm(vap_ynf ~ cig_usef, data=Complete, family="binomial")
  summary(VapingCiglog) #Univariate analysis for binge drinking. 
  
  odds.n.ends(VapingCiglog)
  
VapingBingelog <- glm(vap_ynf ~ bingedrinkf, data=Complete, family="binomial")
  summary(VapingBingelog) #Univariate analysis for binge drinking. 
  
  odds.n.ends(VapingBingelog)
  
VapingCoclog <- glm(vap_ynf ~ coke_usef, data=Complete, family="binomial")
  summary(VapingCoclog) #Univariate analysis for cocaine use.
  
  odds.n.ends(VapingCoclog)
  
VapingEcstalog <- glm(vap_ynf ~ ecstasyf, data=Complete, family="binomial")
  summary(VapingEcstalog) #Univariate analysis for ecstasy use.  
  
  odds.n.ends(VapingEcstalog)
  
VapingGenlog <- glm(vap_ynf ~ genderf, data=Complete, family="binomial")
  summary(VapingGenlog) #Univariate analysis for gender.
  
 chisq.test(Complete$genderf, Complete$vap_yn)
  
VapingGradeslog <- glm(vap_ynf ~ gradesf, data=Complete, family="binomial")
  summary(VapingGradeslog) #Univariate analysis for grades.
  
  odds.n.ends(VapingGradeslog)
  
VapingMarilog <- glm(vap_ynf ~ mari_usef, data=Complete, family="binomial")
  summary(VapingMarilog) #Univariate analysis for marijuana use.
  
  odds.n.ends(VapingMarilog)
  
VapingMethlog <- glm(vap_ynf ~ meth_usef, data=Complete, family="binomial")
  summary(VapingMethlog) #Univariate analysis for meth use.

  odds.n.ends(VapingMethlog)
  
VapingHerlog <- glm(vap_ynf ~ heroin_usef, data=Complete, family="binomial")
  summary(VapingHerlog) #Univariate analysis for heroin use.

  odds.n.ends(VapingHerlog)
  
VapingSexlog <- glm(vap_ynf ~ sexf, data=Complete, family="binomial")
  summary(VapingSexlog) #Univariate analysis for sex.
  
 chisq.test(Complete$sexf, Complete$vap_yn)
```


## STEP 11: Logistic Regression wiht complete cases. 

```{r}
#First Model Exposure and Outcome 
Mod0VapCig_use<- glm(vap_ynf ~ cig_usef , data=Complete, family="binomial")
  summary(Mod0VapCig_use)
  
#Second model including sex (not gender) race and age (not grade)
Mod1SexRaceAge<- glm(vap_ynf ~ sexf + racef + agef + cig_usef, data=Complete, family="binomial")
  summary(Mod1SexRaceAge) 

#Logistic analysis including sex, race and grade.
Mod2SexRaceGrade <- glm(vap_ynf ~ sexf + racef + grade_levelf + cig_usef , data=Complete, family="binomial")
  summary(Mod2SexRaceGrade)

odds.n.ends(Mod2SexRaceGrade)

# Compare Model with Age to Model with grade to determine better variable to use
lrtest(Mod1SexRaceAge, Mod2SexRaceGrade)
```


## Run Logistic Regression with complete dataset: FINAL MODEL
```{r}
CompleteCaseMod<-glm(vap_ynf ~ sexf + racef + grade_levelf + genderf + bingedrinkf + mari_usef + coke_usef + ecstasyf + meth_usef + gradesf + heroin_usef + cig_usef, family="binomial", data= Complete)
summary(CompleteCaseMod)

# Get ORs and CIs
odds.n.ends(CompleteCaseMod)

# Get Log Likelihood
logLik(CompleteCaseMod)

# Compare Final Model to Model with just Cigarette smoking and Vaping
lrtest(Mod0VapCig_use, CompleteCaseMod)
```


## STEP 12: Run Logistic Regression with Imputed data 

```{r}
AdjustLogImp <-with(imp, glm(vap_ynf ~ sexf + racef + grade_levelf + genderf + bingedrinkf + mari_usef + coke_usef + ecstasyf + meth_usef + gradesf + heroin_usef + cig_usef, family="binomial" ))#get regression coefficients
pool(AdjustLogImp) #pool regression coefficients and standard errors 

summary(pool(AdjustLogImp))

ImpSum<-summary(pool(AdjustLogImp)) #Created a variable of the pool imputed data
```


## STEP 13: Calculate odds ratio of imputed data. 

```{r}
round(summary(pool(AdjustLogImp), conf.int = TRUE, exponentiate = TRUE), 3)
```

## Make sure to compare ORs for the Imputed Dataset to those for the Complete Dataset

```{r}
# ORs do not differ much.
```


## STEP 14: Determine TOP 5 influential observations (using Complete Dataset) using a Cook's Distance plot

```{r}
plot(CompleteCaseMod, which=4, id.n=5, col="red", cex.id=0.60) 


help(plot)

#identify observations with a Cook's D greater than 0.0004 
y<-as.data.frame(cooks.distance(CompleteCaseMod))
colnames(y)[1]<-"CD"
y$obs_no<-rownames(y)
z<-y[which(y$CD>0.0004),]
z$obs_no
```


## STEP 15: Exclude the top 5 influential observations and compare Betas between models with and without these observations.

```{r}
#car library needed for compareCoefs (notice the Camelcase!)
#dropping obs with CD>0.0004
CompleCase.modex <- update(CompleteCaseMod,subset=c(-1907, -2668, 
                                                               -3873, -3140, 
                                                               -4517))
#compare coefficients between models with and without influential observations, 
#caveat model number of observations is not the same
compareCoefs(CompleteCaseMod, CompleCase.modex) 

#Interpretation: after removing the infucential cases the coefficients did not change much. So we can leave them in.
```



